Draft Final Report: Measles risk assessment, modelling, and benefit--cost analyses for New Zealand

### 2015-04-24 draftfinalreport_v3.Rnw
updates cost benefit

Though the global incidence of measles is declining, in recent years that decline has slowed (\autoref{fig:trendincidence}) and immigration rates to New Zealand have risen (\autoref{fig:travel}). This suggests that the risk of measles importation could increase, though further analyses are require to understand the interaction between these variables.  Of note, however, is the clear seasonality in immigration and travel (\autoref{fig:travel}).  This seasonality suggests that there may be period of increased risk of measles importation, though again the interactions with seasonal measles transmission from the nations of origin will be an important factor in determining the risk of measles importation. Of interest is the asynchrony, or lack of phase, between peak non-New Zealander immigration and New Zealanders travelling. The Christmas summer period being an exception to this, when there is substantial New Zealander travel and combined these lead to a very sharp peak in travel rates in December (\autoref{fig:travel}).

@@

The estimated $R_v$ for each outbreak is shown in \autoref{fig:r0}. Our analyses suggest the $R_v$ for the 2009--2014 outbreaks range between 0.18--3.92, and for the last outbreak that ended in August 2014, but for which data were available until 12 June 2014, a mean estimate of 3.66 (range 3.4--3.92). The probability density of the $R_v$ estimates for 13 of 20 outbreaks for which we had outbreak numbers include one. Of particular note there have been a number of outbreaks for which our estimated $R_v$ was well above one and thus we may expect outbreaks to occur if conditions remain the same. An important caveat to the analysis of the last outbreak is that because this outbreak that ended in August was ongoing when the data were collected, and not in decline, $R_v$ is necessarily over one, and so the comparison with others must be cautious. However, this is not the only case for which the $R_v$ is greater than one (\autoref{fig:r0}).

@@

To use the results from our modelling exercise to help inform the appropriate measles vaccination coverage, we use \autoref{eq:prop}. The proportion of the population requiring additional vaccination to make $R_v$ < 1 ranged from 17\% to 34\% at the DHB level, with a national average of approximately 28\% (\autoref{table:attack}). These additional vaccination numbers can be calculated in a number of different ways, as discussed in the benefit--cost section (\autoref{sub:cost_benefit}). However, they require differing numbers of vaccinations per DHB. Estimates for West Coast, Wairarapa, and South Canterbury, for example, are fewer than 1000 (\autoref{table:attack}). The estimated numbers to vaccinate in the Auckland area, however, are higher for each DHB, with those estimated for Waitemata, Auckland, and Counties Manukau DHBs all over 17,000  (\autoref{table:attack}). Note that there is also evidence for vaccine failure (\autoref{fig:vaccstat}) and pockets of susceptible populations \citep{glass4} may require greater vaccination effort than these estimates and this is considered further in the \autoref{sec:cb}.

@@

About 50 years ago, approximately 135 million cases of measles causing 7--8 million deaths were believed to occur in the world \citep{clements4}. Thirty years later, it was estimated there were still approximately 45 million cases of measles occurring annually, including 6 million measles-related fatalities. It was estimated that in 1999 measles was responsible for more than 30 million disability adjusted life years (DALYs) lost and 12 million in 2005 \citep{wolfson7}. The incidence of cases was reduced by more than 50\% from 43 million in 1999 to approximately 20 million in 2005. Approximately 7.5 million deaths from measles were avoided from 2000--05 due vaccination \citep{wolfson7}. The World Health Organization (WHO) estimated 158,000 deaths from approximately 355,000 measles cases in 2011 \citep{who13}.  In addition to the substantial losses occurring in measles-endemic countries, a significant impact is felt in heavily measles-vaccinated countries, which may be considered measles-free, due to contact with cases either in the country of origin or in the previously measles-free country.

The annual cost of treating and controlling measles in 11 industrialised countries was estimated to be more than US\$150 million\footnote{Jan 2001 \$ to US\$ rate 0.4445: \$337.5million} \citep{carabin3}. The estimated cost for a case ranged from US\$189--344\footnote{Jan 2001 \$ to US\$ rate 0.4445: \$425--774} \citep{carabin3}; however, the average estimated cost of a typical hospital case ranges from US\$967--1,755\footnote{Jan 2001 \$ to US\$ rate 0.4445: \$2175--3948} \citep{carabin2}. \citep{stack11} estimated the economic benefits from cases averted due to measles vaccination. They estimated that the expanded vaccination from 2005 to 2015 in 72 of the world's poorest countries could result in nearly US\$10 billion\footnote{Jan 2009 \$ to US\$ rate 0.5526: \$18.1billion} of costs averted between 2011 and 2020. Ninety-nine percent of these averted costs were the result of lost productivity due to an estimated 360,000 measles-specific premature mortalities, with the remaining <1\% associated with averted treatment costs and reduced caretaker productivity for the nearly 12 million measles cases avoided.

@@

Wages lost due to measles are calculated for the period January 2008 - August 2014. We assume that the losses to the economy are the same, whether the the employee is able to obtain sick leave pay, as this is the general loss to the economy as work days are lost. Calculations are based on the assumptions that 5 days of work are lost for each case; however, individuals between 0-14 years of age are not assumed to be employed and therefore did not suffer an income loss and the employment rate of teenagers 15-19 years of age was 41.9\%  (Statistics New Zealand, 2013). If the case is < 20 years of age, it was assumed there was an income loss of 5 days for the care giver, in addition to the wage loss of the case if 15-19 years of age.  Case contact data was used to estimate further wage loss costs. These data were provided by the Ministry of Health. The data included the number and duration of school exclusions.

@@

The total cost for the 293 publicly funded hospital discharges with a measles primary diagnosis that spent 470 nights in hospital from 2001--2014 (part year) was \$550,024 (\autoref{table:hosp}). The mean cost per case was \$1,877. The mean cost per day of stay in the hospital was \$1,170.

@@

Regression analyses showed statistically significant associations between cost of hospitalisation and three variables: length of hospitalisation, case age and year of case, and a less strong association with case gender (\autoref{table:regression_costs}). Results showed the expected hospitalisation costs in 2000 of a female measles patient who did not stay overnight in the hospital was \$582.  The cost was \$256 less if the case was a male. It increased of approximately \$406 per night of hospitalisation and \$64 per year over the time period of 2000--2014. The cost of a case decreased with the age of the patient by approximately \$8 per year of case age.

@@

Wages lost for the period January 2008 -- August 2014 for the 247 cases and care givers was estimated to be \$207,155. This consisted of \$104,539 for the cases and \$102,616 for the care giver, but did not include wage losses for cases under 15 years of age. Overall, the cost per case from 2008--2014 was estimated to be \$2923.27 (\$839 in forgone wages, \$1,765 in management costs, and \$1,710 in hospital costs). This figure brings an approximate estimate of \$546650.78 for 187 cases for the current outbreak in 2014 alone, which is comprised of earnings lost, case management and hospitalisation costs for 17\%.

@@

The cost figures above are used to estimate the savings for vaccinating additional populations, using estimated per case costs saved. The costs of the catch up vaccination schemes are estimated to be two different values, to determine how sensitive the benefit--cost ratio (B/C) was to differing vaccination costs. Values of \$20 and \$50 per vaccine are used, based on US literature, however, this assumes all costs are included with the vaccine delivery. Therefore, we also estimate approximate cost a vaccine (including the delivery costs) would need to be before the vaccination costs exceed the benefits. Thus the expected measles-related costs due to constant introduction of measles despite increased population immunity and the costs of the vaccination schemes are used to estimate the costs of additional vaccine programs for the B/C analysis. The financial savings from reduced measles cases are used to work out the benefits for the additional vaccination programs. The costs for the catch up vaccination programs are presumed to be over a single year. The benefits for the cases saved are presumed to be over a ten year period. Because benefits are assessed over a 10--year time period, a discounting rate of 3\% discount per year for the costs saved was used, as is common for healthcare discounting \citep{honeycutt6}. Measles introductions are expected to be annually occurring events, though the costs per year for each event was also discounted. Thus, the B/C ratio was:

@@

For the cost analyses we used the values from the above cost section (\autoref{sub:cost_benefit}). Specifically, we used the average cost of a case for the analyses to be \$839 lost in wages, \$1765.49 in case management costs, and \$1877 in hospitalisation costs for those attending hospital, with 17\% of cases predicted to be hospitalised (\autoref{table:hosp}). We estimated there would be approximately one introduction of measles per year (\autoref{sub:risk_analyses}). We provide two costs for measles vaccinations for our cost analyses, \$20 and \$50, based on US literature. We add GP consultation costs of \$20 and the cost of contact exclusions at \$2624.57 (see \autoref{sub:cost_res}). Note vaccination and GP costs can be altered by the Ministry of Health in the spreadsheet provided.

@@

It is worth noting that vaccination strategies that target the very young (<1 year old) may be less effective, as our analyses of the vaccinated cases suggests a substantial proportion of vaccinated cases that were vaccinated (\autoref{fig:ageandvac}) were vaccinated with a single vaccine at a very young age (\autoref{fig:vaccstat}). Furthermore, it may be unnecessary to vaccinate very young, as it appears possible to reach the appropriate figure (28\% of currently na\"{\i}ve) by vaccinating all the currently na\"{\i}ve 2--17 year olds, so pre-school and school age children. However, there are a number of different vaccination schemes that will lead to sufficient coverage and these are given in \autoref{table:coh_vac}. The appropriate cover could be reached by catch up vaccination of all school aged children.

@@

\subsection{Benefit--cost analysis summary}
\begin{itemize}
\item The mean wage losses per measles case is estimated to be \$839
\item The mean cost per measles case attending hospital is estimated to be \$1,877
\item Approximately 17\% of measles cases attend hospital
\item The mean management cost per measles case for the current outbreak, estimated from three months of data is estimated to be \$1,765
\item Using $R_0$ values estimated for measles in New Zealand prior to vaccination and the current na\"{\i}ve population sizes, the benefits of catch up vaccination strategies are clear (>1 B/C ratio).
\item Introduction of measles are estimated by simulation to lead to median outbreak size of 2 cases, suggesting most infections will be individuals or lead to single secondary cases. However, mean sizes of 61 cases we predicted, despite $R_v$ being below one and the epidemic predicted to die out without additional interventions. There is, however, a small chance for some of these outbreaks to reach larger sizes.
\end{itemize}

>>>>

Though the global incidence of measles is declining, in recent years that decline has slowed (\autoref{fig:trendincidence}) and immigration rates to New Zealand have risen (\autoref{fig:travel}). This suggests that the risk of measles importation could stay at similar levels or increase, though further analyses are require to understand the interaction between these variables.  Of note, however, is the clear seasonality in immigration and travel (\autoref{fig:travel}).  This seasonality suggests that there may be period of increased risk of measles importation, though again the interactions with seasonal measles transmission from the nations of origin will be an important factor in determining the risk of measles importation. Of interest is the asynchrony, or lack of phase, between peak non-New Zealander immigration and New Zealanders travelling. The Christmas summer period being an exception to this, when there is substantial New Zealander travel and combined these lead to a very sharp peak in travel rates in December (\autoref{fig:travel}).

@@

The estimated $R_v$ for each outbreak is shown in \autoref{fig:r0}. Our analyses suggest the $R_v$ for the 2009--2014 outbreaks range between 0.18--3.92, and for the last outbreak that ended in August 2014, but for which data were available until 12 June 2014, a mean estimate of 3.66 (range 3.4--3.92). The probability density of the $R_v$ estimates for 13 of 20 outbreaks for which we had outbreak numbers include one. Of particular note there have been a number of outbreaks for which our estimated $R_v$ was well above one and thus we may expect outbreaks to occur if conditions remain the same. An important caveat to the analysis of the last outbreak is that because this outbreak that ended in August was ongoing when the data were collected (12 June 2014), and not in decline, $R_v$ is necessarily over one, and so the comparison with others must be cautious. However, this is not the only case for which the $R_v$ is greater than one (\autoref{fig:r0}).

@@

To use the results from our modelling exercise to help inform the appropriate measles vaccination coverage, we use \autoref{eq:prop}. The proportion of the na\"{i}ve population requiring additional vaccination to make $R_v$ < 1 ranged from 17\% to 34\% at the DHB level, with a national average of approximately 28\% (\autoref{table:attack}). These additional vaccination numbers can be calculated in a number of different ways, as discussed in the benefit--cost section (\autoref{sub:cost_benefit}). However, they require differing numbers of vaccinations per DHB. Estimates for West Coast, Wairarapa, and South Canterbury, for example, are fewer than 1000 (\autoref{table:attack}). The estimated numbers to vaccinate in the Auckland area, however, are higher for each DHB, with those estimated for Waitemata, Auckland, and Counties Manukau DHBs all over 17,000  (\autoref{table:attack}). Note that there is also evidence for vaccine failure (\autoref{fig:vaccstat}) and pockets of susceptible populations \citep{glass4} may require greater vaccination effort than these estimates and this is considered further in the \autoref{sec:cb}.

@@

About 50 years ago, approximately 135 million cases of measles causing 7--8 million deaths were believed to occur in the world \citep{clements4}. Thirty years later, it was estimated there were still approximately 45 million cases of measles occurring annually, including 6 million measles-related fatalities. It was estimated that in 1999 measles was responsible for more than 30 million disability adjusted life years (DALYs) lost and 12 million in 2005 \citep{wolfson7}. The incidence of cases was reduced by more than 50\% from 43 million in 1999 to approximately 20 million in 2005. Approximately 7.5 million deaths from measles were avoided from 2000--05 due to vaccination \citep{wolfson7}. The World Health Organization (WHO) estimated 158,000 deaths from approximately 355,000 measles cases in 2011 \citep{who13}.  In addition to the substantial losses occurring in measles-endemic countries, a significant impact is felt in heavily measles-vaccinated countries, which may be considered measles-free, due to contact with cases either in the country of origin or in the previously measles-free country.

The annual cost of treating and controlling measles in 11 industrialised countries was estimated to be more than US\$150 million\footnote{Jan 2001 \$ to US\$ rate 0.4445: \$337.5million} \citep{carabin3}. The estimated cost for a case ranged from US\$189--344\footnote{Jan 2001 \$ to US\$ rate 0.4445: \$425--774} \citep{carabin3}; however, the average estimated cost of a typical hospital case ranges from US\$967--1,755\footnote{Jan 2001 \$ to US\$ rate 0.4445: \$2175--3948} \citep{carabin2}. \citet{stack11} estimated the economic benefits from cases averted due to measles vaccination. They estimated that the expanded vaccination from 2005 to 2015 in 72 of the world's poorest countries could result in nearly US\$10 billion\footnote{Jan 2009 \$ to US\$ rate 0.5526: \$18.1billion} of costs averted between 2011 and 2020. Ninety-nine percent of these averted costs were the result of lost productivity due to an estimated 360,000 measles-specific premature mortalities, with the remaining <1\% associated with averted treatment costs and reduced caretaker productivity for the nearly 12 million measles cases avoided.

@@

Wages lost due to measles are calculated for the period January~2008--August~2014. We assume that the losses to the economy are the same, whether the the employee is able to obtain sick leave pay, as this is the general loss to the economy as work days are lost. Calculations are based on the assumptions that 5 days of work are lost for each case; however, individuals between 0--14 years of age are not assumed to be employed and therefore did not suffer an income loss and the employment rate of teenagers 15--19 years of age was 41.9\%  (Statistics New Zealand, 2013). If the case is < 20 years of age, it was assumed there was an income loss of 5 days for the care giver, in addition to the wage loss of the case if 15--19 years of age.  Case contact data was used to estimate further wage loss costs. These data were provided by the Ministry of Health. The data included the number and duration of school exclusions.

@@

The total cost for the 293 publicly funded hospital discharges with a measles primary diagnosis that spent 470 nights in hospital from 2000--2014 (part year) was \$550,024 (\autoref{table:hosp}). The mean cost per hospitalised case was \$1,877. The mean cost per day of stay in the hospital was \$1,170.

@@

Regression analyses showed statistically significant associations between cost of hospitalisation and three variables: length of hospitalisation and year of case, and a less strong association with case age and case gender (\autoref{table:regression_costs}). Results showed the expected hospitalisation costs in 2000 of a female measles patient who did not stay overnight in the hospital was \$582.  The cost was \$256 less if the case was a male. It increased approximately \$406 per night of hospitalisation and \$64 per year over the time period of 2000--2014. The cost of a case decreased with the age of the patient by approximately \$8 per year of case age.

@@

Wages lost for the period January~2008--August~2014 for the 247 cases and care givers was estimated to be \$207,155. This consisted of \$104,539 for the cases and \$102,616 for the care giver, but did not include wage losses for cases under 15 years of age. Overall, the cost per case from 2008--2014 was estimated to be \$2923.27 (\$839 in forgone wages, \$1,765 in management costs, and \$1,877 in hospital costs for the 17\% of cases hospitalised). This figure brings an approximate estimate of \$546650.78 for 187 cases for the current outbreak in 2014 alone, which is comprised of earnings lost, case management and hospitalisation costs.

@@

The cost figures above are used to estimate the savings for vaccinating additional populations, using estimated per case costs saved. The costs of the catch up vaccination schemes are estimated to be two different values, to determine how sensitive the benefit--cost ratio (B/C) was to differing vaccination costs. Values of \$20 and \$50 per vaccine are used, based on US literature, however, this assumes all costs are included with the vaccine delivery. Therefore, we also estimate approximate cost a vaccine (including the delivery costs) would need to be before the vaccination costs exceed the benefits. Thus the expected measles-related costs due to constant introduction of measles despite increased population immunity and the costs of the vaccination schemes are used to estimate the costs of additional vaccine programs for the B/C analysis. The financial savings from reduced measles cases are used to estimate the benefits for the additional vaccination programs. The costs for the catch up vaccination programs are presumed to be over a single year. The benefits for the cases saved are presumed to be over a ten year period. Because benefits are assessed over a 10--year time period, a discounting rate of 3\% discount per year for the costs saved was used, as is common for healthcare discounting \citep{honeycutt6}. Measles introductions are expected to be annually occurring events, though the costs per year for each event was also discounted. Thus, the B/C ratio was:

@@

For the cost analyses we used the values from the above cost section (\autoref{sub:cost_benefit}). Specifically, we used the average cost of a case for the analyses to be \$839 lost in wages, \$1765.49 in case management costs, and \$1877 in hospitalisation costs for those attending hospital, with 17\% of cases predicted to be hospitalised (\autoref{table:hosp}). We estimated there would be approximately one introduction of measles per year (\autoref{sub:risk_analyses}). We provide two costs for measles vaccinations for our cost analyses, \$20 and \$50, based on US literature. We add GP consultation costs of \$20 and the cost of contact exclusions at \$2624.57 (see \autoref{sub:cost_res}). Note vaccination and GP costs can be altered by the Ministry of Health in the spreadsheet provided. In addition, laboratory costs which were not included above may also be altered in the spreadsheet.

@@

In corcordance with current measles knowledge, strategies that target the very young (<1 year old) may be less effective, as our analyses of the vaccinated cases suggests a substantial proportion of vaccinated cases that were vaccinated (\autoref{fig:ageandvac}) were vaccinated with a single vaccine at a very young age (\autoref{fig:vaccstat}). Furthermore, it may be unnecessary to vaccinate very young, as it appears possible to reach the appropriate figure (28\% of currently na\"{\i}ve) by vaccinating all the currently na\"{\i}ve 2--17 year olds, so pre-school and school age children. However, there are a number of different vaccination schemes that will lead to sufficient coverage and these are given in \autoref{table:coh_vac}. The appropriate cover could be reached by catch up vaccination of all school aged children.

@@

\subsection{Benefit--cost analysis summary}
\begin{itemize}
\item The mean wage losses per measles case is estimated to be \$839.
\item The mean cost for measles cases that attend hospital is estimated to be \$1,877.
\item Approximately 17\% of measles cases attend hospital.
\item The mean management cost per measles case for the current outbreak, estimated from three months of data is estimated to be \$1,765.
\item Using $R_0$ values estimated for measles in New Zealand prior to vaccination and the current na\"{\i}ve population sizes, the benefits of catch up vaccination strategies are clear (>1 B/C ratio).
\item Introduction of measles are estimated by simulation to lead to median outbreak size of 2 cases, suggesting most infections will be individuals or lead to single secondary cases. However, mean sizes of 61 cases we predicted, despite $R_v$ being below one and the epidemic predicted to die out without additional interventions. There is, however, a small chance for some of these outbreaks to reach larger sizes.
\end{itemize}
